---
title: "Week 2 Assignment: R Markdown and Leaflet"
subtitle: "JHU Data Science | Developing Data Products"
date: "2023-01-06"
output: html_document
---

Create a web page using R Markdown that features a map created with Leaflet. 

## Setup
```{r, echo=TRUE, message=FALSE}
library(dplyr)
library(leaflet)
library(lubridate)
```

## Import and Update the Data
Here we import a CSV file containing the data. The data contains the date and locations of upcoming tests of a dog activity/sport ([NACSW scent detection](https://nacsw.net/)), along with a URL of the test host's website where the premium/test entry information can be found.
```{r, echo=TRUE, comment="", collapse=TRUE}
df <- read.csv("forleaflet.csv")
str(df)

df <- df %>% 
  mutate(
    label = paste(location, " on ", date, sep = ""),
    site = paste("<a href='", url, "'> Link to Premium </a>", sep = ""), 
    wday = wday(df$date, label=TRUE))
```

## Create a Leaflet map
Using leaflet, we'll create a map of the test locations. At zoomed out levels, the locations will cluster into a single markers. At the zoomed in level, the individual location markers will be blue if the test is on a weekend, red if the test is on a weekday. Hovering over a marker will reveal the location and test date associated with the marker. Clicking on a marker will reveal a clickable link to the test host's website for test entry.
```{r, echo=TRUE, out.width=910}
getColor <- function(df) {
  sapply(df$wday, function(wday) {
    if(wday == "Fri" || wday == "Sat" || wday == "Sun") {
      "blue"
    } else {
      "red"
    } }
  )
}

icons <- awesomeIcons(
  icon = 'ios-paw',
  iconColor = 'white',
  library = 'ion',
  markerColor = getColor(df)
)

df %>%
  leaflet() %>%
  addTiles() %>%
  addAwesomeMarkers(
    ~lng, ~lat, icon=icons, label = ~label, popup =~site, clusterOptions = markerClusterOptions()) %>%
  addLegend("topleft", title = "Test Day", labels = c("Weekday", "Weekend"), colors = c("#cf0c0c", "#03adfc"), opacity = 1)
```